---
title: "On Transmission Types and Mileage"
author: "Riccardo Finotello"
date: "24 June 2020"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

In this report we investigate the impact of gear transmission type on petrol
mileage of several types of cars. The objective is to study what kind of
transmission has the best impact on mileage (measured in miles per gallon) and
quantify the difference between the two types.

## Exploratory Data Analysis

We first load the _mtcars_ dataset and look at its description and summary:

```{r eda.load}
data("mtcars")
str(mtcars)
```
As we can see, the dataset is mainly composed of numerical variables and factors
which can be used for regression.

We will be interested in the `mpg` variable as dependent variable mainly as a
function of the transmission type `am` which is a binary variable describing
_automatic_ ($am = 0$) or _manual_ ($am = 1$) transmission type. In Figure
\ref{fig:eda:box} we show the distribution of the dependent variable (_mpg_) as
a function of the two types of transmission we intend to probe.

In Figure \ref{fig:eda:corr} we show the correlation matrix of the variables
together with their hierarchical structure. There are clear indications of
interdependence between several variables.

## Regression Analysis

We first try, as a base model, to predict the mileage using the most correlated
variables (we do not fit the intercept as, in this context, it has no physical
meaning):

```{r reg.fit1}
fit.corr1 <- lm(mpg ~ cyl + disp + hp + wt + factor(cyl) -1, data = mtcars)
summary(fit.corr1)$coef
```

As we can see most factors are relevant for the fit and we will try to keep them
in further analysis. However the power output of the cars seems to irrelevant.
We further investigate the impact of _hp_ on the model by comparing the results
when we remove it from the fit:

```{r reg.anova}
fit.corr2 <- lm(mpg ~ cyl + disp + wt + factor(cyl) -1, data = mtcars)
anova(fit.corr1, fit.corr2)
```

This shows that the variable does indeed play a role in the model.

We finally investigate the impact of the transmission type on the model (we
intentionally left it out from the previous fits in order to have a good
background model to use for inference):

```{r reg.am}
fit <- lm(mpg ~ cyl + disp + wt + hp + factor(cyl) + factor(am), data = mtcars)
summary(fit)$coef
```

In this case we decide to fit the intercept which represents the base estimate
for automatic ($am = 0$) transmission. With a p-value of 0.9 and
`r fit$df` degrees of freedom we can also say that manual transmission
($am = 1$) has an average better impact on the petrol mileage of
`r round(summary(fit)$coef[7,1], 1)` miles per gallon with a confidence interval of

```{r reg.confint}
confint(fit)[8,]
```

In Figure \ref{fig:reg:res} we finally show the diagnostic plots related to the
fit we performed showing that the distribution of the residuals is indeed
uncorrelated and no outlier issues are present.

## Conclusion

From the analysis it seems that manual transmission cars have a better impact
on petrol mileage which can be quantified in $(`r round(summary(fit)$coef[7,1], 1)` \pm
`r round(summary(fit)$coef[7,2], 1)`)$ miles per gallon.

\clearpage

# Appendix

In this appendix we show most plots and figures associated with the analysis.

The objective of the analysis is the prediction of petrol mileage as a function
of the type of the transmission. We show the bounding boxes of such prediction
as a reference:

```{r eda.box, fig.cap="\\label{fig:eda:box}Boxplot of mileage vs type of transmission"}
boxplot(mpg ~ factor(am, labels = c("automatic", "manual")), data = mtcars,
        xlab = "Transmission type",
        ylab = "Miles per Gallon",
        main = "Mileage per Transmission Type"
       )
```

We can then graphically show the correlation matrix of the dataset.

```{r eda.corr, fig.cap="\\label{fig:eda:corr}Correlation matrix."}
library(lattice)
levelplot(cor(mtcars))
```

We finally show the residual plots related to the analysis:

```{r reg.fit.res, fig.cap="\\label{fig:reg:res}Diagnostic plots of the fit."}
par(mfrow=c(2,2))
plot(fit)
```
