---
title: "Countries of the World"
author: "Riccardo Finotello"
date: "26/06/2020"
output:
    html_document:
        keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset Description

The dataset is available on [Kaggle](https://www.kaggle.com/paultimothymooney/latitude-and-longitude-for-every-country-and-state) and contains data on longitude and latitude of all countries in the world.

```{r download}
dat <- read.csv("world_country_and_usa_states_latitude_and_longitude_values.csv")
```

The dataset is made of `r nrow(dat)` rows and `r ncol(dat)` columns containing
the names of every country in the world and their locations:

```{r show}
library(knitr)
kable(head(dat))
```

We will then be interested only in the columns pertaining the latitude, longitude and country names. We will disregard codes and US related data.

```{r dat}
dat <- dat[,2:4]
```

## Exploratory Data Analysis

We are interested in plotting only complete cases:

```{r complete}
complete <- complete.cases(dat)
```

There are `r sum(!complete)` missing values in the dataset. We will however ignore them for the time being:

```{r complete.exclude}
dat <- dat[complete,]
```

## Data Visualisation

We can then visualise each country on the map:

```{r map}
library(leaflet)
dat[,1:2] %>% leaflet() %>% addTiles() %>% addMarkers(popup=dat$country, clusterOptions=markerClusterOptions())
```