---
title: "Central Limit Theorem and the Exponential Distribution"
author: "Riccardo Finotello"
date: "19 June 2020"
output:
  pdf_document:
    default
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

We consider the exponential distribution and compare it with the Central Limit
Theorem (CLT): we perform several simulations to extract samples and average a
number of them to show that the distribution of the mean is approximately a
normal distribution.

## Simulations

In this simulation we will use exponential distributions with `lambda = 0.2`
(the _rate_ parameter):

```{r par.rate}
lambda <- 0.2
```

We will then define the number of exponentials simulating draws from the
distribution:

```{r par.sim}
n.exp <- 40   # no. of exponentials
n.sim <- 1000 # no. of simulations
```

We then set the seed for reproducible results:

```{r seed}
set.seed(42)
```

We then proceed to the simulation of the distributions:

```{r sim}
means <- NULL
for (i in 1 : n.sim) {means = c(means, mean(rexp(n = n.exp, rate = lambda)))}
```

As a reference we also consider 1000 samples coming from one of the
distributions. Instead of `r print(n.sim)` distributions of `r print(n.exp)`
samples, we consider one distribution with `r print(n.sim)` samples:

```{r dist}
exp.dist <- rexp(n = n.sim, rate = lambda)
```

We then plot the comparison as a preliminary graphic showing the key differences
of the distributions in Figure \ref{fig:diff}.

```{r comp, fig.cap = "\\label{fig:diff}Difference between the distributions of 1000 exponential distribution holding 40 samples and one distribution with 1000 samples.", echo = FALSE}
par(mfrow = c(1,2))
hist(means, xlab = "Means of the distros", main = "Avg. of 1000 distributions")
hist(exp.dist, xlab = "Samples", main = "Exponential distribution")
```

## Sample Mean and Theoretical Mean

We now consider the comparison between the population mean and the expected
value of the estimator for an exponential distribution:

```{r means}
mean.pop <- mean(means)
mean.exp <- 1 / lambda

print(paste("Population mean:", round(mean.pop,3)))
print(paste("Expected mean:", mean.exp))
```

The results reflect the prediction of the CLT: the distribution of the averages
will result in a normal distribution centred in the expected value of the mean.

## Sample Variance and Theoretical Variance

We can then proceed and perform the same analysis for the standard deviation of
the population and the expected estimator:

```{r sd}
sd.pop <- sd(means)
sd.exp <- 1 / lambda

print(paste("Population standard deviation:", round(sd.pop,3)))
print(paste("Exponential standard deviation:", sd.exp))
```

The difference in the values should not be regarded as worrisome: the CLT
ensures that the distribution of the averages tends to a normal distribution
centred in the expected value of the mean and a standard deviation equal to the
standard error (i.e. $\sigma = \frac{s}{\sqrt{n}}$, where $\sigma$ is the
standard error, $s$ is the population standard deviation and $n$ is the number
of degrees of freedom in the distribution):

```{r vars.std}
sd.err <- sd.exp / sqrt(n.exp)

print(paste("Population standard deviation:", round(sd.pop,3)))
print(paste("Standard error:", round(sd.err,3)))
```

Again we then have "experimental" evidence of the validity of the CLT, since the
population mean and the standard error associated to the normal distribution of
the averages are extremely close.

## Distribution

In this last section we provide a visual reference of the compatibility of the
distribution of the averages with a normal distribution. In fact such property
can be immediately visualised in the histogram in Figure \ref{fig:comp}. We
first create the intervals of reference values for the normal distribution.

```{r dist.val}
x <- seq(0.75*min(means), 1.25*max(means), by = 0.01) #---- create the x axis
y <- dnorm(x, mean = mean.exp, sd = sd.err) #-------------- create PDF function
```

```{r dist.plot, fig.cap = "\\label{fig:comp}Comparison between the distribution of the averages and a normal distribution centered in the expected value of the mean estimator (shown in red) and a standard deviation equal to the standard error (shown in blue).", echo = FALSE}
hist(means, prob = TRUE, xlab = "Means of the distros", main = "Avg. of 1000 distributions")
lines(x, y, col = "red")
abline(v = mean.exp, col = "red", lwd = 2) #------------------------ add mean
abline(v = mean.exp + c(-1,1) * sd.err, col = "blue", lwd = 2) #---- add std
arrows(x0 = mean.exp, y0 = 0.3, x1 = mean.exp + sd.err, y1 = 0.3, code = 3, col = "blue")
arrows(x0 = mean.exp, y0 = 0.3, x1 = mean.exp - sd.err, y1 = 0.3, code = 3, col = "blue")
```