---
title: "Brief Analysis of Tooth Growth Data"
author: "Riccardo Finotello"
date: "19 June 2020"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this brief analysis we consider data on the growth of odontoblasts cells in
the teeth of guinea pigs under vitaminic treatment. We provide a short
exploratory data analysis and compare tooth growth by **supp** and **dose**
using basic statistical inference.

## Data

We first load the dataset from the `datasets` package in `R`.

```{r load}
library(data.table)
data(ToothGrowth)

dat <- data.table(ToothGrowth, stringsAsFactors = FALSE)
names(dat) <- c("Length", "Supplement", "Dose")
```

The dataset describes the effects of the dosage of vitamin C or orange juice
(the `Dose` column, numeric) in guinea pigs with different supplements type
(`Supplement`, factor) on the length of teeth growth (the `Length` column,
numeric). We provide a brief summary as a first reference:

```{r summary, echo = FALSE}
str(dat)
```

We will be interested in a twofold analysis: we will investigate the outcome of
administering the vitamins through _orange juice_ (identified by `OJ`) or
_vitamin C_ (`VC`) on the length of the tooth, then we will try to infer
something more on the dosage.

The data is made of both numeric and categorical variables. In particular the
column `Supplement` can be used as a level to distinguish different treatments:

```{r eda.dist}
dat.VC <- dat[Supplement == "VC", .(Length, Dose)] # data for vitamin C
dat.OJ <- dat[Supplement == "OJ", .(Length, Dose)] # data for orange juice
```

```{r eda, echo = FALSE, fig.cap = "\\label{fig:comp}Growth length of odontoblasts for different treatments."}
par(mfrow = c(1,2))
boxplot(Length ~ Dose, data = dat.VC, xlab = "Dose [mg]", main = "Growth with vitamin C")
boxplot(Length ~ Dose, data = dat.OJ, xlab = "Dose [mg]", main = "Growth with orange juice")
```

At a visual level we can already appreciate the difference between different treatments on the length of the odontoblasts.

## Effect of Supplement Type on Tooth Growth

First of all we shall consider the impact of supplement type on the growth. We
first plot the range of variation of the length per supplement type in Figure
\ref{fig:supp}.

```{r boxplot, echo = FALSE, fig.cap = "\\label{fig:supp}Growth of odontoblasts grouped by supplement type."}
boxplot(Length ~ Supplement, data = dat, names = levels(dat$Supplement), ylab = "Length", main = "Growth of odontoblasts per supplement type")
```

It seems that the largest effect on growth is given by giving supplements of
orange juice to the guinea pigs. We can in fact see that the mean are relevantly
higher in this case:

```{r t.test.sup, echo = FALSE}
t.test(Length ~ Supplement, data = dat, paired = FALSE, alternative = "greater")
```

The _t test_ of unpaired, one-sided populations leads to a p-value less than
0.05 which leads to the rejection of the null hypothesis $H_0$ (i.e. there is no
average difference), supported also by the confidence interval of the difference
of the means which is strictly positive: the averange growth using orange juice
seems to be larger than using vitamin C.

## Effect of Dosage of Orange Juice on Tooth Growth

We then explore the impact of the dosage of orange juice on tooth growth:

```{r dose.summary, echo = FALSE}
library(knitr)
kable(summary(dat.OJ))
```

From the plot in Figure \ref{fig:comp} we have already seen the difference in
the outcome when different doses are administered. We then consider the average
growth in length when using the orange juice:
```{r dose.avg, echo = FALSE}
data.OJ.avg <- dat.OJ[, .(Average.Length = mean(Length)), by = Dose]
kable(data.OJ.avg)
```

The question that we ask is therefore whether a higher dosage of orange juice
can lead to larger growths (assuming the null-hypothesis $H_0$ stating the
opposite):

```{r oj.t}
# the test is ordered such that higher doses are the RHS,
# thus we need the "less" alternative
t.1 <- t.test(Length ~ Dose,
              data = dat.OJ[Dose == 0.5 | Dose == 1.0],
              paired = FALSE,
              alternative = "less")
t.2 <- t.test(Length ~ Dose,
              data = dat.OJ[Dose == 0.5 | Dose == 2.0],
              paired = FALSE,
              alternative = "less")
t.3 <- t.test(Length ~ Dose,
              data = dat.OJ[Dose == 1.0 | Dose == 2.0],
              paired = FALSE,
              alternative = "less")

print(paste("P-value between 0.5 mg and 1.0 mg:", t.1$p.value))
print(paste("P-value between 0.5 mg and 2.0 mg:", t.2$p.value))
print(paste("P-value between 1.0 mg and 2.0 mg:", t.3$p.value))
```

It seems therefore that higher doses of orange juice may lead to benefit for the
growth in length of the odontoblasts. We can also look at the confidence
intervals to support the conclusion. In particular between the 95% CI between
0.5 mg and 1.0 mg is `r t.1$conf.int` and between 1.0 mg and 2.0 mg is
`r t.3$conf.int`, which show that they do not include 0.

## Conclusions

From the simple data analysis we can therefore conclude that the growth of the
odontoblasts is heavily influenced by the type of supplement and its dosage. In
particular we showed that the best treatment for a larger growth is the
administration of orange juice to the guinea pigs in larger quantities.