---
title: "Monitoring the Italian Situation on Covid-19"
author: "Riccardo Finotello"
date: "26/06/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Description

### What this is about

- a simple project demonstrating the use of **plot_ly** in `R`
- an insignificant plot showing very simple data

### What you should be aware of

- no recommendation or assumption has been made
- this is unrelated from finding an effective solution to a real problem
- the situation is serious and this is not the place to discuss it

## Data

- data from Italian government available on [GitHub](https://github.com/pcm-dpc/COVID-19)

```{r cache = TRUE}
covid <- read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-andamento-nazionale/dpc-covid19-ita-andamento-nazionale.csv")

# change class of "data"
covid$data <- as.Date(covid$data, "%Y-%m-%dT%H:%M:%S")

# select only relevant columns
covid <- covid[,c(1,4,5)]
names(covid) <- c("date", "intensive", "hospital")

# plot data
library(plotly)
days <- 1:length(covid$date)
covid.plot <- covid %>% plot_ly(x = ~date,
                                y = ~hospital,
                                type="scatter",
                                name="hospitalized",
                                mode="lines+markers",
                                showlegend=TRUE,
                                text=paste(days, "days since lockdown")
                               ) %>% 
                        add_trace(y = ~intensive,
                                  type="scatter",
                                  name="intensive care",
                                  mode="lines+markers",
                                  showlegend=TRUE,
                                  text=paste(days, "days since lockdown")
                                 ) %>%
                        layout(title="Covid-19 Monitoring",
                               yaxis=list(type="log",
                                          title="counts (log scale)"
                                         ),
                               xaxis=list(title="date (lockdown on Feb, 24th)"),
                               hovermode="compare"
                              )
covid.plot
```

## Conclusions

### What we can see

- data show decreasing count of hospitalized people
- intensive care seems to be relieved of some burdain

### Future directions

- we are still **not out of danger**
- precautions should be taken to **avoid a stepback**

